# LITERATURE REVIEW

This chapter focuses on elaborating theories and findings pertaining to resilience and psychological outcomes. @sec-motivation elaborates the motivation behind this research, including the economic evaluation of mental disorders. @sec-attachment summarizes child-parent attachments as proposed by @bowlby1973attachment and @ainsworth1979infant. @sec-resilience explains the scope and context of mental resilience in this research. @sec-neurobio-resilience outlines the underlying neurobiological mechanism in psychological resilience. @sec-stressor defines a set of stressors investigated in this research, which will be the generic precipitators of maladaptive behavior. @sec-psych justifies the psychological outcomes.

```{r init}

pkgs <- c("magrittr")
pkgs_load <- sapply(pkgs, library, character.only = TRUE)

gbd    <- readRDS("../RData/gbd-mental-illness/gbd_describe")
groups <- gbd[[1]]$group %>% set_names(gsub(x = ., "\\W", "_"))

gbd_stats <- lapply(groups, function(grp) {
  lapply(gbd, \(tbl) tbl %>% subset(.$group == grp, select = -group)) %>%
    {do.call(rbind, .)} %>%
    inset2("stat", value = rownames(.)) %>%
    tidyr::pivot_longer(daly_depress:prev_eat_disorder) %>%
    inset(c("metrics", "psych"), value = list(
      "metrics" = gsub(x = .$name, "_.*", ""),
      "psych"   = gsub(x = .$name, ".*_", "")
    ))
})

reportGBD <- function(gbd_stat, m, p) {
  arr <- gbd_stat %>%
    subset(
      subset = {
        .$stat %in% c("mean", "sd") & .$metrics == m & .$psych == p
      },
      select = value
    ) %>%
    extract2(1) %>%
    round(2)

  msg <- "%s%s [%s]"

  if (m == "prev") {
    msg <- sprintf(msg, arr[[1]], "%", arr[[2]])
  } else {
    msg <- sprintf(msg, arr[[1]], "", arr[[2]])
  }

  return(msg)
}

```

## The Scope of Mental Health Burden {#sec-motivation}

Using Global Burden of Diseases (GBD) data from 1990 to 2019, @owidmentalhealth described the scope of problem with mental illnesses, reported as prevalence and disability-adjusted life years (DALY). The GBD dataset enlists 206 countries and 22 country groups according to the World Health Organization, United Nation, and World Bank classification. The clinical diagnoses included in the GBD dataset is depressive disorder, anxiety disorder, bipolar disorder, eating disorder, and schizophrenia. The author replicated this approach in table [-@tbl-summary] - [-@tbl-cor-gbd-4] to elucidate the regional issues based on country-level income grouping. Interestingly, depression is the most prevalent in low-income countries ($\bar{x}$ [$\sigma$]^[$\bar{x}$ $[\sigma]$: mean [standard deviation]] = `r reportGBD(gbd_stats$Low_Income, "prev", "depress")`) and the least prevalent in middle-income country (`r reportGBD(gbd_stats$Middle_Income, "prev", "depress")`). Meanwhile, anxiety is the most prevalent in high-income countries (`r reportGBD(gbd_stats$High_Income, "prev", "anxiety")`) and the least prevalent in lower-middle-income countries (`r reportGBD(gbd_stats$Lower_Middle_Income, "prev", "anxiety")`). The strong and positive association between DALY and prevalence implies the more occurring cases is closely related with increasing burden of diseases. @tbl-cor-gbd-1 - [-@tbl-cor-gbd-4] show that all mental illnesses present with cross association, though with a different pattern occurring in high-level income countries compared to the others. Specifically looking at depression and anxiety, high-income countries presented with a positive association while other countries presented with negative associations. To further address this issue, @Kessler2006 with the Assessment, Classification, and Evaluation Group at the WHO initiated World Mental Health (WMH) survey. WMH complements GBD by providing insight on validated mental health burden worldwide, by incorporating the use of lay-administered Composite International Diagnostic Interview for DSM-IV [@who2004prevalence].

```{r tbl-summary}
#| tbl-cap: Descriptive statistics of mental health indicators from GBD dataset, adapted from @owidmentalhealth

readRDS("../RData/gbd-mental-illness/gbd_report")

```

```{r tbl-cor-gbd-1}
#| tbl-cap: Pairwise correlation of mental illnesses metrics in high-income countries
gbd_cor <- readRDS("../RData/gbd-mental-illness/gbd_cor")
knitr::kable(gbd_cor$high_income)

```

```{r tbl-cor-gbd-2}
#| tbl-cap: Pairwise correlation of mental illnesses metrics in middle-income countries

knitr::kable(gbd_cor$middle_income)

```

```{r tbl-cor-gbd-3}
#| tbl-cap: Pairwise correlation of mental illnesses metrics in lower middle-income countries

knitr::kable(gbd_cor$lower_middle_income)

```

```{r tbl-cor-gbd-4}
#| tbl-cap: Pairwise correlation of mental illnesses metrics in low-income countries

knitr::kable(gbd_cor$low_income)

```

@who2004prevalence highlighted that the combined global prevalence of any mental disorder is ranging from 4.3% to 26.4%, where the most reported cases are anxiety disorder (2.4%-18.2%), mood disorder (0.8%-9.6%), and substance abuse (0.1%-6.4%). While the proportion of severe cases (0.4%-7.7%) in any diagnosis is comparably lower than the prevalence of moderate (0.5%-9.4%) and mild cases (1.8%-9.7%), the impact to productivity is astounding. At least 30 productive days are lost among subjects reporting any severe mental disorder (IQR: 32.1-38.4 days), and accordingly, the number of day loss is lower in subjects with moderate (IQR: 9.2-18.8 days) or mild cases (IQR: 0.1-3.6 days). However, not all cases receives adequate treatment, where the proportion of people receiving treatment is ranging from 0.8% to 15.3% of the affected population. Even though severe cases are more likely to be diagnosed and treated, up to 50.3% of severe cases in developed countries and 85.4% of severe cases in less-developed countries receive no treatment. The proportion of untreated subjects is even higher in moderate and mild cases, ranging from 50%-90.3% and 64.8%-99.5%, respectively. In a follow-up study using WMH data, @Kessler2009 reported that early-onset mental disorders significantly predict a wide range of physical disorders and reduced achievements in a number of role domains. Furthermore, productivity loss due to major depressive disorder in the United States alone is over five weeks of productive-day loss per year, approximated to USD 36 billion of economical loss. Similarly, data from 10 countries participating in WMH survey indicates that an average of 22 productive days loss per worker with attention deficit hyperactivity disorder. Even though the cause of mental disorder varies widely, its impact to the individual and society is apparent.

### Depressive Disorder

This proposal shall outline two facades of depression, i.e. major depressive disorder and depressive symptoms. For brevity, major depressive disorder will be termed as depressive disorder, while the combination of up to four depressive symptoms excluding suicidal ideation will be termed as minor depression. In mental health domain, major depressive disorder is one of the leading cause of disability worldwide. While the prevalence of depressive disorder is only second to anxiety disorder, the impact is immensely dreadful, causing most DALYs for both sexes [@Rehm2019]. Worldwide, depressive disorder has affected over 300 millions people, contributing to 7.5% of years lived with disability. Depressive disorder has a markedly devastating impact from a public health perspective as well, especially taking into account its considerable economic impact. Across the globe, depressive disorder has a lifetime prevalence ranging from 2% in some Asian country to 21% in some European countries [@GutirrezRojas2020]. While depressive disorder is the primary determinant of suicidal behavior and immensely impacting the quality of life, an adequate treatment is not well accessible to the majority of patients [@MorenoAgostino2021].

The hallmark of a depressive disorder is the presence of debilitating depressive episodes occurred at least once and lasting at minimum for two weeks. This episode is usually accompanied by changes in mood, perception on pleasure, and cognitive functions [@Otte2016]. According to DSM-V, people with depressive disorder may present with the following symptoms:

- Depressed mood
- Diminished interest or pleasure in almost all activities
- Weight loss/gain without dieting, accompanied by decrease/increase in appetite
- Sleeping problem, manifesting as insomnia or hypersomnia
- Psychomotor agitation or retardation
- Loss of energy
- Feeling of worthlessness and inappropriate guilt which might be delusional
- Diminished ability to think or concentrate
- Suicidal ideation

Various mechanisms contribute to the development of depressive disorder, including genetic predisposition, brain structure alteration, and changes in the stress-response system [@Otte2016]. Genetic contribution is estimated to explain 35% of depressive disorder cases, with possible heritability as demonstrated in family and twin-based studies, a genetic mechanism which also has an interplay with rare mutations. By investigating the biology of depressive disorder, reduced hippocampal volume is apparent in affected subjects, alongside alteration in cognitive network control and affective-salience network. Furthermore, the presence of depressive disorder also implicates stress-response system, indicated by changes in hypothalamic-pituitary-adrenal axis, the functional state of autonomic nervous system, and the immune system [@Otte2016].

Specific groups in the population is at a higher risk of diagnosed with depressive disorder [@Otte2016]. The most turbulent time in life contributing to the onset of depressive disorder is between mid-to-late adolescence and early 40s, where the prevalence of depressive disorder is gradually peaking with age before early adulthood. The reported median age of onset of depressive disorder is 25 years old in both sexes, with the risk of developing depressive disorder after puberty in women is twice as high as in men [@Otte2016]. The odds of being diagnosed with depressive disorder is 1.4-2.8 times higher in women than in men [@GutirrezRojas2020]. Beside age and sex, depressive disorder is also linked to marital status, i.e. being separated/divorced, but the wide confidence interval of 5.0-74.4 causes the estimated effect size rather indeterminate. Though not completely explanatory, @GutirrezRojas2020 reported other factors contributing to the development of depressive disorder, including employment status, socioeconomic status, education level, physical health, and substance uses. Furthermore, the presence of depressive disorder may also accentuate its association with other mental health-related correlates. Depressive disorder is in particular associated with general anxiety disorder, different classes of phobia, borderline personality disorder, paranoid personality disorder, and neurotic personality trait. Another risk factor to take into account is the history of receiving abusive treatments during childhood, either physical, psychological, or sexual abuse. Each form of abuse has different odds in causing depressive disorder, where adults receiving harsh treatment in early childhood might have 1.6 to 4.1 times of higher odds of diagnosed with depressive disorder [@GutirrezRojas2020].

Physical comorbidities of depressive disorder is also frequently reported, where @Otte2016 listed six somatic correlates of depressive disorder, including heart disease, diabetes mellitus, cancer, cognitive impairment, disability, and obesity. Traumatic experience also dysfunction following physical illness are the cornerstones triggering a depressive episode. Specifically in obesity, distorted body image and loss of self-esteem are highly detrimental to the susceptibility of having a depressive disorder. Moreover, biological factors such as altered inflammatory pathways also implicate the pathogenesis of depression [@GutirrezRojas2020]. Increased circulation of pro-inflammatory cytokines and C-reactive protein also become one of the biological explanation underpinning altered inflammatory pathways in its contribution to the development of depressive disorder.

Depressive disorder and its impact to the patients also poses an economic burden to the society at large. Using WMH survey data from the Netherlands, @Cuijpers2007 estimated the direct and indirect cost of depressive disorder. Similarly reported in other countries, the prevalence of depressive disorder acquired by the WMH survey in the Netherlands was 5.8%. After adjusting to purchasing power parities from OECD, the total economical burden per capita for a year was USD 3,313 (95% CI: USD 1,234-5,390). The calculated associated cost included direct medical cost (USD 478), direct non-medical cost (USD 313), and indirect non-medical cost (USD 2,535).^[The calculation of the economical burden was per Dutch guideline for health economic evaluation. Overall, @Cuijpers2007 provided a thorough and realistic cost estimation. Direct medical costs were defined as health service-related fee to access mental healthcare and medication, including professional fee paid for the general practitioners, social workers, and physiotherapists. Direct non-medical cost were accounted for occurring cost when commuting to access the mental health services, including fee for travel, parking, and time spent when accessing the mental health service. Indirect cost were calculated for productivity loss due to illness.] Even without fulfilling the diagnostic criteria of a major depressive disorder, having minor depression is also associated with a significant economical cost. The resulting annual cost in people with minor depression was USD 2,141 (95% CI: 753-3,529). Even though the direct medical cost (USD 10) and direct non-medical costs (USD 29) were much lower compared to those with depressive disorder, the indirect non-medical cost was within an overlapping range, amounted to USD 2,101 (95% CI: USD 648-3,518) in minor depression and USD 2,535 (95% CI: USD 549-4,520) in depressive disorder [@Cuijpers2007]. Also reported in the USA, the total estimated annual lost due to depressive disorder was USD 16,854 per capita, totalling USD 333.7 billion in 2019 alone. The distribution of cost estimated by @Greenberg2023 is slightly different from @Cuijpers2007 due to the difference in health system, where the total healthcare costs were USD 127.3 billion, household-related costs were USD 80.1 billion, presenteeism was USD 43.3 billion, and absenteeism was USD 38.4 billion. With the steadily increasing prevalence of depressive disorder, it is also apparent that the associated cost is proportionally growing [@Greenberg2021].

### Anxiety Disorder

Anxiety is a quite prevalent symptom being addressed in an outpatient setting, with the associated complaints of fear, nervousness, and panic. The presence of anxiety may precede physical issues presenting in the cardiovascular, gastrointestinal, respiratory, musculoskeletal, or nervous systems [@Martin2003].  The problem with anxiety and nervousness is as ancient as history, where it was first described by @burton1621 in *The Anatomy of Melancholy*.

> "Many lamentable effects this fear causeth in man, as to be red, pale, tremble, sweat; it makes sudden cold and heat come over all the body, palpitation of the heart, syncope, etc. It amazeth many men that are to speak or show themselves in public."
>
> *--- Robert Burton*

Anxiety itself poses a wide array of problems which may affect individuals since an early age of life. In a meta-analysis, @Solmi2021 demonstrated a bimodal distribution of the age of onset for anxiety, obsessive-compulsive, and feeding/eating disorders. The peak ages of onset are 5.5 and 15.5 years old for anxiety and fear-related disorders, 14.5 and 49.5 years old for obsessive-compulsive or related disorders, and 15.5 years old for feeding or eating disorders. Evaluating the peak age of each group, it is sensible that problems related to anxiety may emerge since early childhood and re-emerge during adolescent or late adulthood. After all, these disorders share similar underlying mechanism, i.e. stemming from anxious feelings and nervousness. In this proposal, anxiety disorder is narrowed down to only include anxiety and fear-related disorders, comprising generalized anxiety disorder, social anxiety disorder, and separation anxiety disorder.

Relying on the diagnostic criteria provided by DSM-V, anxiety disorder can be determined in case the presence of excessive worry for at least six months [@DeMartini2019]. Worry and nervousness in an anxiety disorder is expected to be volatile, i.e. one day with less worry than the others, yet it still persists that the number of days feeling worried dominates the rest. The occurring nervousness is quite difficult to control and rationalize, hence impeding the affected individuals from performing day-to-day activities. Worry and nervousness in an anxiety disorder are also associated with at least three of the following symptoms:

- Restlessness or feeling on edge
- Easily fatigued or feeling drained
- Difficulty in concentrating
- Being irritable
- Muscle tension
- Sleep disturbances, which could present as any form of insomnia, i.e. difficulty falling asleep, restless during sleep, or unsatisfying sleep

With the recent WMH survey suggested that the prevalence of anxiety disorder is ranging from 2.4% to 18.2%, it is clear that there are several uncertainties in determining its global prevalence. Previous meta-regression model supported this claim by highlighting different odds ratio in included covariates when determining the heterogeneity of anxiety disorder prevalence. In their meta-regression model, @Baxter2012 fitted covariates that might influence the prevalence of anxiety disorder, grouped by its substantive factors and methodological factors. Identified covariates in the substantive factors are gender, age group, conflict status, culture, economic status, and urbanicity. Meanwhile, identified covariates in the methodological factors are prevalence type, survey instrument, and number of disorders.

Each covariate from the substantive factors contribute to explaining the heterogeneity of anxiety disorder prevalence [@Baxter2012]. Consistent with findings on other mental disorders, women are at a higher risk of being diagnosed with anxiety disorder, with an adjusted OR of 2.1 (95% CI: 1.2-2.3) and predicted global prevalence of 9.9% (95% CI: 6.6%-14.8%). By using 35-54 year-old group as a reference, only older adults (55+ years old) are at a lower risk of having an anxiety disorder, with an adjusted OR of 0.8 (95% CI: 0.7-0.9) and predicted global prevalence of 6.7% (95% CI: 4.4%-10.1%). Younger adults (18-34 years old) and adolescents (3-17 years old) are at the same odds of having anxiety disorder as those aged 35-54 years old, with the predicted global prevalence ranging from 4.7% to 12.5%. The presence of conflict in daily life increases the odds of having anxiety disorder, with predicted global prevalence of 9% (95% CI: 5.9%-13.6%). Using the European and Anglosaxon group as a reference, other cultures are at a lower odds of diagnosed with anxiety disorder. Interestingly, emerging and developed countries are more likely to be affected by anxiety disorder compared to developing country, with a higher odds ratio ranging from 1.1 to 1.9. In contrast, people in urban area are less likely to be affected by anxiety disorder compared to people living in rural area, with a markedly lower predicted global prevalence of 8.4% (95% CI: 5.5%-13%).

Naturally, the heterogeneity of predicted global prevalence is also affected by methodological factors, but only to a lesser degree. The identified contributing covariates to the heterogeneity are the use of past-year prevalence and the presence of multiple disorders. Contributing factors of the heterogeneity is, of course, not to be taken at its face value. These factors do not necessarily define the causality or direct inferential association. Despite uncertain estimates of the global prevalence, the predicted prevalence of 7.3% by @Baxter2012 is highly in line with and confirming findings obtained by @who2004prevalence.

Elaborate:

- Patophysiology
- Cost: Medication, productivity loss

### Burnout

Elaborate:

- Clinical definition
- Symptoms
- Differential diagnosis
- Prevalence
- Cost: Medication, productivity loss

## Attachment Theory {#sec-attachment}

@bowlby1973attachment, @hinde1996attachment, @bretherton2013origins

## Resilience as an Indicator of Mental Immunity \label{resilience} {#sec-resilience}

Resilience can be seen as a trait or a state of a human psyche, where
resilience as a trait covers contributing internal factors that help the
individual to overcome adversities. As a state, resilience is best defined as a
dynamic adaptive behaviour that emerges due to circumstantial changes. Several
constituents of resilience as a trait includes habitual effective coping
mechanism and positive emotions, both act as a buffer against negative life
events [@davydov2010resilience]. Meanwhile, as a state, resilience is a dynamic
capacity to adapt supported by the interaction of internal and external
factors, including biological, psychological, social, and ecological systems
[@ungar2020resilience]. Resilience as a trait is observable as a non-adaptive
countermeasures, where the individual will act against the stressful events,
i.e. being resistant to adversities. On the other hand, resilience as state is
an adaptive transformation to prevent persistent dysfunction or
psychopathology [@davydov2010resilience], thus, will be better suited for any
modelling and policy-directed research.

The intertwining facade of the trait and state of resilience is explainable as
an inherent activities of neurocircuitries [@rutten2013resilience], where
gene-environmental interactions play a role in determining individual
susceptibility to psychiatric illness (resilience as a trait)
[@amstadter2014psychiatric]. Also, there exists biological mechanism that allow
experience-dependant plasticity and fuel up adaptability (resilience as a
state) [@derakhshan2020attentional; @rutten2013resilience]. By considering both
sides of resilience, This research is concerned with the latter aspect, where
maintaining an adaptive behavior is of essence to abate the development of
mental disorder.

## The Neurobiology of Resilience {#sec-neurobio-resilience}

@vella2019theoretical, @rutten2013resilience

@vella2019theoretical & @liu2018biological

- Resilience as a dynamic process regulating homeostatic plasticity


## Stressors Precipitating the Development of Mental Issues \label{stressor} {#sec-stressor}

Resilience is the moderator between risk factors and psychological disorders, therefore, a clear definition of included stressors will aid in developing an explainable and generalizable study design [@ungar2020resilience]. Defining the stressors also allow us to measure resilience using a reactivity approach, where resilience is evaluated as growing tolerance to regular stress [@davydov2010resilience]. This way we can partly assess the dynamic process in developing an adaptive behavior.

The presence of stressors disrupts physiological function to a lower utility state [@oken2015systems]. Following the methods proposed by @veer2021psycho, our research evaluates two distinct groups of stressor, namely the general daily stressor and health-related stressor. The general stressor is an instrument comprising ten items of daily hassles and life events, while the health-related stressor consists of items reflecting adversity related to general health concern in self, families, and close relatives. Any item rated as `"TRUE"` is then weighted for the severity, ranging from *not burdensome* (0) to *substantially burdensome* (5). Below is the excerpt of the general stressor items:

- Negative political events
- Conflicts with relatives or colleagues
- Financial problems
- Physical health problems
- Mental health problems
- Burdensome experiences with family
- Burdensome experiences in workplace
- Conflicts with strangers
- Separation from a loved one, which triggers reduced social contact and
  loneliness
- Death of a loved one

The health-related stressor in @veer2021psycho was particularly narrating distress caused by the COVID-19 pandemic. In brief, these items are comprised of trepidation or perceived burden due to a disease, thus adaptable for a more general research context.

## Psychological Outcomes \label{psych} {#sec-psych}

Depression, anxiety, and burnout are selected as outcomes of interest due to
the relatedness between burnout $\rightleftharpoons$ depression and burnout
$\rightleftharpoons$ anxiety [@koutsimani2019relationship]. The partial
correlations have a moderate effect size, signifying the boundary between
burnout and depression also burnout and anxiety. Therefore, statistically,
each of depression, anxiety, and burnout does not share similar latent
construct even though they have closely-related manifests. This is
interpretable as distinctive remarks to differentiate three of such outcomes
and the clinical importance of examining the others when one is present.

Burnout is a condition of increasing depersonalization and emotional
exhaustion, leading to reduced personal accomplishment in the workplace (sense
of inefficacy). Burnout often happens due to a long-term ongoing stressor,
where the inability to cope with external issues decreases work engagement
[@Schaufeli2017]. The incidence of burnout overlaps with psychological
distress, marked by depressive symptoms or anxiety.

In a recent finding, burnout is partially explainable by depression and
anxiety. Depression symptoms pose similar characteristics to exhaustion and
depersonalization, while anxiety is related to the sense of personal inefficacy
[@golonka2019occupational]. In theory, depression is marked by the loss of
energy (fatigue) and anxiety is marked by being excessively worried; but a
study suggested the presence of anxious depression [@gaspersz2018patients],
which in turn also implies a pathophysiological overlap between depression and
anxiety.  Moreover, lower brain connectivity in the default mode network is a
curious co-occurrence between depression [@jacob2020neural], anxiety
[@imperatori2019default], and burnout [@shang2022aberrant]. Neurobiological,
clinical, and contextual overlaps between depression, anxiety, and burnout
necessitate the inclusion of these outcomes to infer a more general impact of
associated stressors.

Several policies have been implemented to alleviate the mental health burden in
Indonesia, mainly highlighted in the Mental Health Act in 2014
[@tampubolon2021covid], and further defined in the Indonesian Minister of
Health Regulation Number 77 of 2015 and Indonesian Minister of Health
Regulation Number 54 of 2017 [@hartono2018tracking]. During the pandemic, the
government of Indonesia has stipulated a Presidential Decree R1 No. 7 of 2020
to promptly respond the pandemic; however, the practical implementation has not
addressed mental health issues. Though, several ongoing initiatives led by the
government has been realized as a regulation on managing a digital counselling
platform [@tampubolon2021covid]. The digital platform, conducted through
various chat-based and video call-based application, helps the general
population to get in touch with therapists, psychologists, counselors, and
educational practitioners [@ifdil2020online]. Tele-consulting, either through
chat or video call, enables practitioners to provide psychological first aid to
address elevated distress during the pandemic, depression, and anxiety
[@ifdil2020online].

:::{.content-hidden unless-profile="site"}

# References {-}

:::
